---
title: "Week 2 Exercises"
author: "Can"
date: "2026-09-10"
output: html_document
---
# header 1
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 4:

```{r Ex 4 A}
# Load required package
library(bootUR)
library(ggplot2)
library(dplyr)
library(vars)

# Load dataset
data(MacroTS)

# get the GDP and CONS data from the MacroTS set


# take only the relevant time frame
gdp_uk <- window(MacroTS[, "GDP_UK"], start = c(1995, 1), end = c(2019, 4))
cons_uk <- window(MacroTS[, "CONS_UK"], start = c(1995, 1), end = c(2019, 4))

#sapply(MacroTS, class) 

# plot them 
#ggplot(MacroTS, aes(x = 1:nrow(MacroTS), y = `GDP_UK`)) +
#  geom_line() +
#  geom_point()
#ggplot(MacroTS, aes(x = 1:nrow(MacroTS), y = `CONS_UK`)) +
#  geom_line() +
#  geom_point()

plot(gdp_uk)
plot(cons_uk)

```
We can see that both graphs have a clear upward trend with one negative shock around halfway down the time frame. It's clearly not stationary, as the mean changes over time.
The shock is also clearly during the 2008 crisis.

Now we want to take the log-differences. For this I'll first take the log of both sets and then take the differences

```{r Ex 4 b}
# take the logs
lgdp <- log(gdp_uk)
lcons <- log(cons_uk)

# take the differences
dgdp <- diff(lgdp)
dcons <- diff(lcons)

# now we make a var with these two, I think that means we just put them together and make a vector
y <- cbind(dgdp,dcons)

# to select order use VARselect which shows what lag orders look like 
lags <- VARselect(y)

# to keep it "parsimonious" and avoid overfitting i'll use BIC (called SC apparently)
print(lags$selection)
print(lags$criteria)

```
```{r Ex 4 c}
# use a function from the VARS library to estimate the VAR with order 1
var <- VAR(y, p = 1, type = "const")

# check the estimated output
summary(var)

```
From the summary, we can see that

We now want to check the Granger causality to see if consumption growth causes gdp growth

```{r Ex 4 d}
# we can use the causality function to chekc this
causality(var, cause="dcons")
```
The result suggests that there is causality between the two as the p-value below the confidence interval, meaning that we reject the null hypothesis that consumption growth doesn't cause gdp growth.

I'll apply the impulse for finding the IRF to the model instead of a specific variable (gdp or cons)
```{r Ex 4 e}
# use the irf function
irf <- irf(var)
plot(irf)
```
Difference between the upper and lower impulse responses????
The impulse response on gdp in the var shows that 

f) No :)
I would say that using in-built features (functions) is never ideal, as we are not directly privy to the inner workings of the functions. So it could be improved by doing all the calculations manually. 
Another point is


# header 2

Exercise 5:

My two favourite ARMA models are AR(2) and ARMA(1,1), so I will be using them. I'll use the arima() function to implement them as I feel that Enrico was encouraging that (and I do not have any phi coefficients in mind).
The time series is 100 units long, so the initial sample size will be s = 40 (10 years of data), and will iteratively increase until h = 24 (the last 5 years).

```{r}
# already have the uk gdp as a variable
ssize = 40
n = length(dgdp) - ssize
wn <- rnorm(length(dgdp), 0, sd)

# initialise the error vectors of the models
errModel1 <- numeric(n) 
errModel2 <- numeric(n)

# so, we want to start with a sample size of 40, go up to 76
for (i in 1:n) {
  # set our sample for this iteration, subtract 1 as index i must start at 1
  ysample <- dgdp[1:(ssize+i-1)]
  
  # now we get our ARMA models for this sample
  fitm1 = arima(ysample, order = c(2,0,0))
  fitm2 = arima(ysample, order = c(1,0,1))
  
  # now we do the ARMA forecasts
  # first we do the AR(2) one
  mu1 <-  coef(fitm1)["intercept"]
  phi11 <- coef(fitm1)["ar1"]
  phi12<- coef(fitm1)["ar2"]
  err1 <-  wn[ssize+i]
  fore1 <-  mu1 + phi11*ysample[ssize+i] + phi12*ysample[ssize]
  
  # now the ARMA(1,1)
  mu2 <- coef(fitm2)["intercept"]
  phi2   <- coef(fitm2)["ar1"]
  theta2 <- coef(fitm2)["ma1"]
  err21 <-  wn[ssize+i]
  err22 <- wn[ssize]
  fore2 <-  mu2 + phi2*ysample[ssize+i] + theta2*err22 + err21
  
  # finally we add the errors to the forecast error vectors we made
  errModel1[i] <- ysample[ssize+i] - fore1
  errModel2[i] <- ysample[ssize+i] - fore2
}

# now we need the MSFE
msfe1 <-  mean(errModel1^2)
msfe2 <- mean(errModel2^2)

```

