---
title: "Week 2 Exercises"
author: "Can"
date: "2026-09-10"
output: html_document
---
# header 1
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise 4:

```{r Ex 4 A}
# Load required package
library(bootUR)
library(ggplot2)
library(dplyr)
library(vars)

# Load dataset
data(MacroTS)

# get the GDP and CONS data from the MacroTS set


# take only the relevant time frame
gdp_uk <- window(MacroTS[, "GDP_UK"], start = c(1995, 1), end = c(2019, 4))
cons_uk <- window(MacroTS[, "CONS_UK"], start = c(1995, 1), end = c(2019, 4))

#sapply(MacroTS, class) 

# plot them 
#ggplot(MacroTS, aes(x = 1:nrow(MacroTS), y = `GDP_UK`)) +
#  geom_line() +
#  geom_point()
#ggplot(MacroTS, aes(x = 1:nrow(MacroTS), y = `CONS_UK`)) +
#  geom_line() +
#  geom_point()

plot(gdp_uk)
plot(cons_uk)

```
We can see that both graphs have a clear upward trend with one negative shock around halfway down the time frame. It's clearly not stationary, as the mean changes over time.
The shock is also clearly during the 2008 crisis.

Now we want to take the log-differences. For this I'll first take the log of both sets and then take the differences

```{r Ex 4 b}
# take the logs
lgdp <- log(gdp_uk)
lcons <- log(cons_uk)

# take the differences
dgdp <- diff(lgdp)
dcons <- diff(lcons)

# now we make a var with these two, I think that means we just put them together and make a vector
y <- cbind(dgdp,dcons)

# to select order use VARselect which shows what lag orders look like 
lags <- VARselect(y)

# to keep it "parsimonious" and avoid overfitting i'll use BIC (called SC apparently)
print(lags$selection)
print(lags$criteria)

```
```{r Ex 4 c}
# use a function from the VARS library to estimate the VAR with order 1
var <- VAR(y, p = 1, type = "const")

# check the estimated output
summary(var)

```
From the summary, we can see that

We now want to check the Granger causality to see if consumption growth causes gdp growth

```{r Ex 4 d}
# we can use the causality function to chekc this
causality(var, cause="dcons")
```
The result suggests that there is causality between the two as the p-value below the confidence interval, meaning that we reject the null hypothesis that consumption growth doesn't cause gdp growth.

I'll apply the impulse for finding the IRF to the model instead of a specific variable (gdp or cons)
```{r Ex 4 e}
# use the irf function
irf <- irf(var,runs=1313,ortho=TRUE)
plot(irf)
```
Difference between the upper and lower impulse responses????
The impulse response on gdp in the var shows that 

f) No :)
I would say that using in-built features (functions) is never ideal, as we are not directly privy to the inner workings of the functions. So it could be improved by doing all the calculations manually. 
Another point is


# header 2

Exercise 5:

My two favourite ARMA models are AR(2) and ARMA(1,1), so I will be using them. I'll use the arima() function to implement them as I feel that Enrico was encouraging that (and I do not have any phi coefficients in mind).
The time series is 100 units long, so the initial sample size will be s = 40 (10 years of data), and will iteratively increase until h = 24 (the last 5 years).

```{r EX 5 a}
# already have the uk gdp as a variable
# ssize = 40
# n = length(dgdp) - ssize
# wn <- rnorm(length(dgdp), 0, sd)
# 
# # initialise the error vectors of the models
# errModel1 <- numeric(n) 
# errModel2 <- numeric(n)
# 
# # so, we want to start with a sample size of 40, go up to 76
# for (i in 1:n) {
#   # set our sample for this iteration, subtract 1 as index i must start at 1
#   ysample <- dgdp[1:(ssize+i-1)]
#   
#   # now we get our ARMA models for this sample
#   fitm1 = arima(ysample, order = c(2,0,0))
#   fitm2 = arima(ysample, order = c(1,0,1))
#   
#   # now we do the ARMA forecasts
#   # first we do the AR(2) one
#   mu1 <-  coef(fitm1)["intercept"]
#   phi11 <- coef(fitm1)["ar1"]
#   phi12<- coef(fitm1)["ar2"]
#   err1 <-  wn[ssize+i]
#   fore1 <-  mu1 + phi11*ysample[ssize+i] + phi12*ysample[ssize]
#   
#   # now the ARMA(1,1)
#   mu2    <- coef(fitm2)["intercept"]
#   phi2   <- coef(fitm2)["ar1"]
#   theta2 <- coef(fitm2)["ma1"]
#   last_res2 <- tail(residuals(fitm2), 1) # filtered residual eps_t
#   pred2  <- mu2 + phi2 * (tail(ysample, 1) - mu2) + theta2 * last_res2
#   
#   # finally we add the errors to the forecast error vectors we made
#   errModel1[i] <- ysample[ssize+i] - fore1
#   errModel2[i] <- ysample[ssize+i] - fore2
# }

T_total <- length(dgdp)

# 2. Expanding window parameters
# Use first 40 quarters (10 years) as initial estimation sample
T_start <- 40 
n_forecasts <- T_total - T_start

forecast_errors_m1 <- numeric(n_forecasts)
forecast_errors_m2 <- numeric(n_forecasts)
# 3. Expanding window loop
for (i in 1:n_forecasts) {
  t_end <- T_start + i - 1
  
  # Step i: Define estimation sample (1 to t_end)
  y_train <- y[1:t_end]
  actual  <- y[t_end + 1]
  
  # Step ii: Estimate ARMA models
  fit1 <- arima(y_train, order = c(2, 0, 0), include.mean = TRUE)
  fit2 <- arima(y_train, order = c(1, 0, 1), include.mean = TRUE)
  
  # Step iii: Manual 1-step-ahead forecasts
  # AR(2): y_{t+1|t} - mu = phi * (y_t - mu) phi1 * (y_t-1 - mu)
  mu1  <- coef(fit1)["intercept"]
  phi11 <- coef(fit1)["ar1"]
  phi12 <- coef(fit1)["ar2"]
  pred1 <- mu1 + phi11 * (tail(y_train, 1) - mu1) + phi12 * (tail(y_train, 2)- mu1)
  
  # ARMA(1,1): y_{t+1|t} - mu = phi * (y_t - mu) + theta * eps_t
  mu2    <- coef(fit2)["intercept"]
  phi2   <- coef(fit2)["ar1"]
  theta2 <- coef(fit2)["ma1"]
  last_res2 <- tail(residuals(fit2), 1) # filtered residual eps_t
  pred2  <- mu2 + phi2 * (tail(y_train, 1) - mu2) + theta2 * last_res2
  
  # Step iv: Store forecast errors
  forecast_errors_m1[i] <- actual - pred1
  forecast_errors_m2[i] <- actual - pred2
}

# 4. Compare Mean Squared Forecast Errors (MSFE)
msfe1 <- mean(forecast_errors_m1^2)
msfe2 <- mean(forecast_errors_m2^2)
msfe1
msfe2
plot(forecast_errors_m1^2, type = "l", col = "blue", ylab = "Squared forecast error",
     main = "Squared forecast errors over time")
lines(forecast_errors_m2^2, col = "red")
```
We see that the MSFE for AR(2) is 1.154 and ARMA(1,1) is 1.142. From this we can see that the ARMA(1,1) model is a better forecaster out-of-sample for this specific dataset; however, the less complex AR(2) model is not far behind.

```{r Ex 5 b}
# to adjust we need a window size, I'll take 40 again
w <- 40

for (i in 1:n_forecasts) {
  start <- i
  end <- i + w - 1
  y_train <- y[start:end]
  actual <- y[end + 1]
  
  # Step ii: Estimate ARMA models
  fit1 <- arima(y_train, order = c(2, 0, 0), include.mean = TRUE)
  fit2 <- arima(y_train, order = c(1, 0, 1), include.mean = TRUE)
  
  # Step iii: Manual 1-step-ahead forecasts
  # AR(2): y_{t+1|t} - mu = phi * (y_t - mu) phi1 * (y_t-1 - mu)
  mu1  <- coef(fit1)["intercept"]
  phi11 <- coef(fit1)["ar1"]
  phi12 <- coef(fit1)["ar2"]
  pred1 <- mu1 + phi11 * (tail(y_train, 1) - mu1) + phi12 * (tail(y_train, 2)- mu1)
  
  # ARMA(1,1): y_{t+1|t} - mu = phi * (y_t - mu) + theta * eps_t
  mu2    <- coef(fit2)["intercept"]
  phi2   <- coef(fit2)["ar1"]
  theta2 <- coef(fit2)["ma1"]
  last_res2 <- tail(residuals(fit2), 1) # filtered residual eps_t
  pred2  <- mu2 + phi2 * (tail(y_train, 1) - mu2) + theta2 * last_res2
  
  # Step iv: Store forecast errors
  forecast_errors_m1[i] <- actual - pred1
  forecast_errors_m2[i] <- actual - pred2
}

msfe1 <- mean(forecast_errors_m1^2)
msfe2 <- mean(forecast_errors_m2^2)
msfe1
msfe2
plot(forecast_errors_m1^2, type = "l", col = "blue", ylab = "Squared forecast error",
     main = "Squared forecast errors over time")
lines(forecast_errors_m2^2, col = "red")
```
Here we see that the MSFE for AR(2) becomes 1.168 and ARMA(1,1) becomes 1.145. Both seem to be worse forecasters when compared to the expanding window simulation.

For an h forecast we just need to let a horizon h be defined and then do the same process as before.

```{r Ex 5 c}
# define h
h <- 4

# take an adjusted number of forecasts 
n_forecasts <- T_total - T_start - h + 1
forecast_errors <- numeric(n_forecasts)
w <- 40

# in this case the sampling window is still the rolling version
for (i in 1:n_forecasts) {
  start <- i
  end <- i + w - 1
  y_train <- y[start:end]
  actual <- y[end + h]
  
  # fit the model again
  fit <- arima(y_train, order = c(1, 0, 0), include.mean = TRUE)

  # in the case of an AR(1) (for simplicity) we get:
  mu1  <- coef(fit)["intercept"]
  phi1 <- coef(fit)["ar1"]
  pred_h <- mu1 + (phi1^h) * (tail(y_train, 1) - mu1)  
  
  # store the forecast again
  forecast_errors[i] <- actual - pred_h
}
# then take msfe for this forecast
msfe <- mean(forecast_errors^2)
msfe
plot(forecast_errors^2, type = "l", col = "blue", ylab = "Squared forecast error",
     main = "Squared forecast errors over time")
```
This gives us the general forecast horizon version.

Part d) 

