---
title: "Week 6"
author: "Can"
date: "2026-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 6 Exercises

Exercises for week 6 TSA.

4.
A)

```{r}
# first sset seed so all simulations are reproducible
source("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/code/ssm_functions.R")
set.seed(69)

# then do the initialisations
n <- 200
var_eps <- 1
var_eta <- 0.1
mu1 <- 0

# now we need to generate our y_t and mu_t series and the epsilon values
mu <- numeric(n)
y <- numeric(n)
eps <- rnorm(n = 201, mean = 0, sd = sqrt(var_eps))
eta <- rnorm(n = 201, mean = 0, sd = sqrt(var_eta))

# now we need to use the mu1 to start the data simulation
mu[1] <- mu1
y[1] <- mu1 + eps[1]

for(i in 2:n) { 
  mu[i] <- mu[i-1] + eta[i]
  y[i] <- mu[i] + eps[i]
}

plot(mu)
plot(y)

```
B)
Want to use the functions given by Tobias to get filtered and smoothed estimates.

```{r}
# write model parameters as matrices
model <- list(
  T = matrix(1, nrow = 1,1),
  R = matrix(1, nrow = 1,1),
  Z = matrix(1, nrow = 1,1),
  Q = matrix(var_eta, nrow = 1),
  H = matrix(var_eps, nrow = 1),
  #c = matrix(0, nrow = 1,1),
  a1 = matrix(mu1, nrow = 1,1),
  P1 = matrix(1e7, nrow = 1,1) # from lecture slides, to diffuse shock(?)
)

# Kalman filter needs a matrix
ymat <- matrix(y, ncol = 1)

# use the filter function
kal <- kalman_filter(y = ymat, mod = model)

# now smoothing function
smo <- state_smoother(ymat, mod, kal)

# now we need to take 
at <- kal$a[1:n]
att <- kal$att
ahat <- smo$ahat

par(mfrow=c(1,3))
plot(at, type = "l")
plot(att, type = "l")
plot(ahat, type = "l")

```
We have the point wise estimates.
Now we want the 90% confidence bands. We wil get them from the Pt|t value for the Kalman filter and the Vt value for the smoothing.

```{r}
# get variances first
var_kal <- kal$Ptt[1, 1, ]
band_kal <- sqrt(var_kal)

var_smo <- smo$V[1, 1, ]
band_smo <- sqrt(var_smo)

# now we plot the the estimates with their respective terms
# first the kalman filter
plot(y, col="grey", type="l")
lines(mu, col="deeppink")
lines(att, col="purple", lwd=2)
lines(att + 1.645 * band_kal, col="indianred", lty=5, lwd=2)
lines(att - 1.645 * band_kal, col="indianred", lty=5, lwd=2)

# now the smoothing
plot(y, col="grey", type="l")
lines(mu, col="deeppink")
lines(ahat, col="pink", lwd=2)
lines(ahat + 1.645 * band_smo, col="maroon", lty="dashed", lwd=2)
lines(ahat - 1.645 * band_smo, col="maroon", lty="dashed", lwd=2)

```
Now we need to compute the RMSE of at att and ahat relative to mut
```{r}

rmse_at <- sqrt(mean((mu-at)^2))
rmse_att <- sqrt(mean((mu-att)^2))
rmse_ahat <- sqrt(mean((mu-ahat)^2))
rmse <- data.frame(
  Term = c(
    "at",
    "att",
    "ahat"
  ),
  
  RMSE = c(
    rmse_at,
    rmse_att,
    rmse_ahat
  )
)
rmse
```
The rmse for ahat is the smallest. The reason for this would be that it is the estimate conditioned on the whole dataset of observed $y_t$ values, meaning it has the best understanding of the "layout" of the data. 
The a_tt is in the middle as it is the estimated value accounting for the variation from the observed data, meaning it gets "pulled" closer to the real value observed (in order to get a better prediction).
a_t performs the worst as it is a pure estimate of the data.

5.
We first want to load the data, and transform it to suit our conditions

```{r}
# load data
dss <- read.csv("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/data/GDPC1.csv")

# keep data from 1948Q1 onward
gdp <- data.frame(quarter = as.Date(dss$observation_date, format = "%Y-%m-%d"), gdp = dss$GDPC1)

gdp <- gdp[gdp$quarter >= as.Date("1948-01-01"), ]
gdp
y <- 100*log(gdp$gdp)

# now set the covid values
y[gdp$quarter %in% c("2020-04-01", "2020-07-01")] <- NA
y

```
Now we have the data, we can start.

A)
Now we want to estimate the model and report the estimates. So we first need to instantiate teh model like in question 4, but have to slightly adjust based on lecture slides page 20/12.
Looking at Tobias' code, we see we need to make a function to make the model.

```{r}
# have the model already
source("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/code/ssm_functions.R")
source("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/code/ar2_cycle_simulation.R")
source("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/code/uc_univariate.R")
source("C:/Users/Can/Desktop/eor/tsa/week 6 data+code/code/run_all.R")

run_all

```
Get output from run_all.R script, now pass these thetas into our model.
```{r}
# take the values manually
phi1 <- 1.5133  
phi2 <- -0.5648
delta0 <- 0.9845 
delta1 <- -0.3230
var_eta <- 0.3107 
var_kap <- 0.3286 
  
ymat <- matrix(y, nrow=1)

T = matrix(c(1, 0, 0,
             0, phi1, phi2,
             0, 1, 0), nrow = 3,3)
R = matrix(c(1, 0,
             0, 1,
             0, 0), nrow = 3,2)
Z = matrix(c(1, 1, 0), nrow = 1)
Q = matrix(c(var_eta, 0,
             0, var_kap), nrow = 2, 2)
H = matrix(0, nrow = 1)

# know delta t is d_0 + d_1 past 1972Q4 (3 x y matrix), so manually input these
# see index of 1973Q1 is from 5 to 105
deltat <- matrix(0, nrow = 3, length(y))
deltat[1,1:100] <- delta0
deltat[1,101:length(y)] <- delta0+delta1
c = matrix(c(deltat,
             0,
             0), nrow = 3,1)
a1 = matrix(mu1, nrow = 1,1)
P1 = matrix(1e7, nrow = 1,1) # from lecture slides, to diffuse shock(?)

model <- list(Z=Z, H=H, T=T, R=R, Q=Q, c=c, 
              a1=c(y[1], 0, 0), P1=diag(c(1e7, 1e7, 1e7)))

# now try the kalman filter to estimate teh model
kf2 <- kalman_filter(ymat,model)

```

